---
title: "An example workflow for building rangeModelMetadata objects"
author: "Jamie Kass, Cory Merow, Brian Maitner, Hannah Owens, Brian Enquist, Rob Guralnick"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{An example workflow for building rangeModelMetadata objects}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = FALSE)

```

```{r,results='hide'}
library(rangeModelMetadata)
library(spocc)
library(raster)
library(dismo)
library(ENMeval)
library(dplyr)
library(sp)
library(spThin)
```

# Get data
```{r}
# search GBIF for occurrence data
bv <- spocc::occ('Bradypus variegatus', 'gbif', limit=300, has_coords=TRUE)
# restrict table to coordinates
occs <- dplyr::select(bv$gbif$data$Bradypus_variegatus, longitude, latitude)
# remove duplicate values
occs <- occs[!duplicated(occs),]
# get environmental rasters from dismo package folder
files <- list.files(path=paste(system.file(package='dismo'), '/ex',
                               sep=''), pattern='grd', full.names=TRUE)
# make a stack of the rasters
envs <- raster::stack(files)
```

# Prepare data

```{r}
# thin points to remove spatial autocorrelation
occs.thinned <- spThin::thin.algorithm(data.frame(occs), thin.par=20, reps=1)
# make a spatial points object for the occurrences
occs.sp <- sp::SpatialPoints(occs.thinned[[1]])
# get the bounding box
bb <- sp::bbox(occs.sp)
# make extent object and buffer by 5 degrees
bb.buf <- raster::extent(bb[1]-5, bb[3]+5, bb[2]-5, bb[4]+5)
# crop environmental rasters by the extent to make the background extent
bg.ext <- raster::crop(envs, bb.buf)
# generate up to 10,000 random points within the background extent (we'll sample from the "biome" raster as it has more holes)
bg <- dismo::randomPoints(bg.ext[[9]], n=10000) 
```

# Modeling

```{r}
# run ENMeval
e <- ENMeval::ENMevaluate(occs, bg.ext, bg, method='block', RMvalues=1:3, fc=c('L','LQ'), algorithm='maxent.jar')
```

# Model performance
```{r}
# find the index number of the model with the highest mean test AUC: this will be the optimal model
i <- which(e@results$avg.test.AUC  == max(e@results$avg.test.AUC ))
# extract the optimal model object
m <- e@models[[i]]
```

# Make predictions
```{r}
# create a cloglog prediction and plot
#p <- predict(m, bg.ext, args=c("outputformat=cloglog"))
notNA <- complete.cases(values(bg.ext))
tmp <- predict(m, values(bg.ext)[notNA,], type='link')
p <- bg.ext[[1]] 
values(p) <- NA 
values(p)[notNA] <- tmp
plot(p)
# further south extent for model transfer
t.bb <- raster::extent(-80,-40,-60,-30)
# crop environmental predictors by transfer extent
t.ext <- raster::crop(envs, t.bb)
```
# Transfer model
```{r}
# transfer the model to this new area in cloglog form and plot
notNA.t <- complete.cases(values(t.ext))
#p.t <- predict(m, t.ext, args=c("outputformat=cloglog"))
tmp <- predict(m, values(t.ext)[notNA.t,], type='link')
p.t <- t.ext[[1]] 
values(p.t) <- NA 
values(p.t)[notNA.t] <- tmp
plot(p.t)
```

# Build an `rmm` object
We start by filling in some key authorship information.
```{r}
# generate an empty range model metadata template
rmm <- rmmTemplate(family=c('base')) 
rmm$authorship$rmmName <- 'MerowMaitnerOwensKassEnquistGuralnick_2018_Acer_Maxent_b3'
rmm$authorship$names <- 'Merow, Cory and Maitner, Brian and Owens, Hannah and Kass, Jamie and Enquist, Brian and Guralnick, Rob'
rmm$authorship$contact <- 'cory.merow@gmail.com'
rmm$authorship$relatedReferences <- '@article{, title={RMMS: Species’ Range Model Metadata Standards },author={Merow, Cory and Maitner, Brian and Owens, Hannah and Kass, Jamie and Enquist, Brian and Guralnick, Rob},journal={Unknown},year={2018}, publisher={Unknown}}'
```
<!-- Would be nice to show how to make this a formal citation class, but i don't knowhow: -->
<!-- ref='@article{, title={RMMS: Species’ Range Model Metadata Standards },author={Merow, Cory and Maitner, Brian and Owens, Hannah and Kass, Jamie and Enquist, Brian and Guralnick, Rob},journal={Unknown},year={2018}, publisher={Unknown}}' -->
<!-- as(ref,'citation') -->


To make it easier to fill some `rmm` fields, we provide autofill functions that extract relevant information from common R objects used in the 

```{r}
# occurrence data
rmm <- rmmAutofillspocc(rmm, bv$gbif) 
# autofill info for environmental rasters used for model training and transfer
rmm <- rmmAutofillEnvironment(rmm, bg.ext, transfer=0) 
rmm <- rmmAutofillEnvironment(rmm, t.ext, transfer=1)
# autofill for ENMeval
rmm <- rmmAutofillENMeval(rmm, e, 
                       selectionCriteria="highest mean test AUC",
                       optimalModelIndex=i) 
# R packages used
rmm <- rmmAutofillPackageCitation(rmm=rmm,
                              packages=c('spocc','sp','raster','dismo','ENMeval'))
```  

Now we need to fill in a few fields manually. 

```{r}
rmm$dataPrep$geographic$spatialThin$rule <- "20km used as max distance between points"
# fill in remaining fields for model prediction in background extent
rmm$output$prediction$units <- "relative occurrence rate"
rmm$output$prediction$minVal <- cellStats(p, min)
rmm$output$prediction$maxVal <- cellStats(p, max)
# clamping is the default for predict()
rmm$output$prediction$extrapolation <- "clamping"
```

Let's see how we're doing so far by printing all the filled fields.
```{r}
#print just the non-NULL fields
rmmCleanNULLs(rmm)
```

Now we'll add in some more key entities that aren't readily autfilled
```{r}
rmm$data$occurrence$yearMin <- 1970
rmm$data$occurrence$yearMax <- 2000
```


Check the final object by printing all the filled fields.
```{r}
#print just the non-NULL fields
rmmCleanNULLs(rmm)
# you can also use this function to omit the NULLs at the end of your workflow using, so if you're happy with the above...
rmm <- rmmCleanNULLs(rmm)
```

Now run all availables checks to be sure you're ready to go.
```{r}
rmmCheckFinalize(rmm)
```



